---
title: "Lab 11"
author: "Andrew Graves"
date: "November 20/22 2019"
output:
  ioslides_presentation:
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.height = 4)
```

# Parametric statistical tests as linear models

## Materials fully borrowed from: 

*Common statistical tests are linear models (or: how to teach stats)*

- Full credit to Jonas Kristoffer Lindelov for this post, although these statistical ideas are not new.

**GitHub Post**

- https://lindeloev.github.io/tests-as-linear/

**Cheat Sheet**

- https://lindeloev.github.io/tests-as-linear/linear_tests_cheat_sheet.pdf

## The classic iris dataset

```{r iris}
library(tidyverse)
theme_set(theme_bw())
theme_update(text = element_text(family = "serif"), 
axis.title.y = element_text(margin = margin(r = 20)))

data("iris")
str(iris)
```

## Statistical questions

**One sample t-test**

- Is versicolor petal width greater than 1.2 cm?

**Independent samples t-test**

- Is there a difference in sepal width between setosa and virginica?

**Paired samples t-test**

- Are the sepals longer than the petals?

**One-way ANOVA**

- Is there a difference in sepal length across the three species?

**For all the following plots, blue lines are intercepts and red lines are slopes!**

# One sample t-test

## Is versicolor petal width greater than 1.2 cm?

```{r one_sample_t1}
one_samp_data <- iris %>%
  filter(Species == "versicolor") %>%
  select(Species, Petal.Width)
```

## One sample t.test function

```{r one_sample_t2}
t.test(one_samp_data$Petal.Width, mu = 1.2)
```

## Subtract comparison value from output vector for lm equivalent

```{r one_sample_t3}
one_samp_lm <- lm(one_samp_data$Petal.Width - 1.2 ~ 1)
summary(one_samp_lm)
```

## Plot the one sample model

```{r plot1}
one_samp_plot <- one_samp_data %>%
  ggplot(aes(x = Species, y = Petal.Width)) + 
  geom_jitter(height = .1, width = 0, alpha = 0.5) +
  geom_hline(yintercept = 1.2, lty = 2) + 
  geom_segment(aes(x = .75, xend = 1.25, 
              y = mean(one_samp_data$Petal.Width), 
              yend = mean(one_samp_data$Petal.Width)), color = "blue")
```

## Print the one sample plot

```{r one_samp_plot}
print(one_samp_plot)
```

# Independent samples t-test

## Is there a difference in sepal width between setosa and virginica?

```{r ind_sample_t1}
ind_samp_data <- iris %>%
  filter(Species %in% c("setosa", "virginica")) %>%
  select(Species, Sepal.Width)

ind_samp_data_set <- ind_samp_data %>%
  filter(Species == "setosa") %>%
  select(Sepal.Width) %>%
  pull()

ind_samp_data_virg <- ind_samp_data %>%
  filter(Species == "virginica") %>%
  select(Sepal.Width) %>%
  pull()
```

## Independent samples t.test function

```{r ind_sample_t2}
t.test(ind_samp_data_virg, ind_samp_data_set, var.equal = TRUE)
```

## Assume equal variance across groups for lm equivalent

```{r ind_sample_t3}
ind_samp_lm <- lm(Sepal.Width ~ Species, data = ind_samp_data)
summary(ind_samp_lm)
```

## Plot the independent samples model

```{r plot2}
ind_samp_plot <- ind_samp_data %>%
  ggplot(aes(x = Species, y = Sepal.Width, group = 1)) + 
  geom_jitter(height = .1, width = 0, alpha = 0.5) +
  stat_summary(fun.y = mean, geom = "line", color = "red") +
  geom_segment(aes(x = .75, xend = 1.25, y = mean(ind_samp_data_set), 
               yend = mean(ind_samp_data_set)), color = "blue")
```

## Print the independent samples plot

```{r ind_samp_plot}
print(ind_samp_plot)
```

# Paired samples t-test

## Are the sepals longer than the petals?

- Paired samples t.test function

```{r pair_sample_t2}
t.test(iris$Sepal.Length, iris$Petal.Length, paired = TRUE)
```

## Subtract one vector from the other as output for lm equivalent

```{r pair_sample_t3}
pair_samp_lm <- lm(Sepal.Length - Petal.Length ~ 1, data = iris)
summary(pair_samp_lm)
```

## Plot paired data

```{r plot3}
paired_plot <- iris %>%
  select(Sepal.Length, Petal.Length, Species) %>%
  mutate(id = factor(1:nrow(iris))) %>%
  gather(key = `Flower Attribute`, value = Measurement, -id, - Species) %>%
  ggplot(aes(x = `Flower Attribute`, y = Measurement, group = id, color = id)) + 
  geom_line() +
  geom_jitter(height = .1, width = 0, alpha = 0.5) +
  facet_grid(.~Species) + 
  theme(legend.position = "none")
```

## Print the paired plot

```{r paired_plot}
print(paired_plot)
```

## Plot the paired samples model

```{r plot4}
add_length_diff <- iris %>%
  mutate(`Length Difference` = Sepal.Length - Petal.Length)

paired_samp_plot <- add_length_diff %>%
  ggplot(aes(x = Species, y = `Length Difference`)) + 
  geom_jitter(height = .1, width = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, lty = 2) + 
 geom_segment(aes(x = .75, xend = 3.25, y = mean(add_length_diff$`Length Difference`), 
 yend = mean(add_length_diff$`Length Difference`)), color = "blue")
```

## Print the paired-samples plot

```{r paired_samp_plot}
print(paired_samp_plot)
```

# One-way ANOVA

## Is there a difference in sepal length across the three species?

```{r anova_1}
group_means <- iris %>%
  group_by(Species) %>%
  summarize(sep_length = mean(Sepal.Length))

set_mean <- group_means %>%
  filter(Species == "setosa") %>%
  select(sep_length) %>%
  pull()

vers_mean <- group_means %>%
  filter(Species == "versicolor") %>%
  select(sep_length) %>%
  pull()

virg_mean <- group_means %>%
  filter(Species == "virginica") %>%
  select(sep_length) %>%
  pull()
```

## One-way ANOVA aov function

```{r anova_2}
anova_aov <- aov(Sepal.Length ~ Species, data = iris)
summary(anova_aov)
```

## Setup is similar for lm equivalent

```{r anova_3}
anova_lm <- lm(Sepal.Length ~ Species, data = iris)
summary(anova_lm)
```

## Plot the one-way ANOVA model (w/ treatment contrasts!)

```{r plot5}
anova_plot <- iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) + 
  geom_jitter(height = .1, width = 0, alpha = 0.5) +
  geom_segment(aes(x = .75, xend = 3.25, y = set_mean, yend = set_mean), color = "blue") + 
  geom_segment(aes(x = 1, xend = 2, y = set_mean, yend = vers_mean), color = "red") +
  geom_segment(aes(x = 2, xend = 3, y = set_mean, yend = virg_mean), color = "red")
```

## Print the one-way ANOVA plot (w/ treatment contrasts!)

```{r anova_plot}
print(anova_plot)
```

# Now it's your turn!

Download the in-class activity and answer the questions as best you can. 

Collaborate with your peers to write code. 

I am available for programming questions.