---
title: "Lab 8"
author: "Andrew Graves"
date: "October 30/November 1, 2019"
output:
  ioslides_presentation:
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.height = 4)
```

# Permutation testing

## Permutations

**Definition**

- Random sequences of lists

- Can contain numeric or string values

- Samples without replacement

- The order of the samples matter

**Testing**

- Typically involves randomization of indices.

- Generates null distributions of parameters and test statistics. 

- Does not make parametric assumptions about test statistics. Instead, empirical null distributions of test statistics are compared to observed outcomes. 

- Resampling allows estimating uncertainty.

## Permutation examples

```{r permutations}
(num_vec <- 1:5)
(string_vec <- LETTERS[1:5])

set.seed(42) # Random permutation of numbers 1 to 5
(num_perm <- sample(num_vec, replace = FALSE)) 
set.seed(42) # Random permutation of letters A to E
(string_perm <- sample(string_vec, replace = FALSE))
```

## Subsetting on permutations

```{r subset}
num_vec[num_perm]
string_vec[num_perm]

# The following two expressions produce NAs, becuase you cannot subset on 
# string values
num_vec[string_perm]
string_vec[string_perm]
```

## Set up simulation with correlated data

```{r set_up, message = FALSE, warning = FALSE}
# Modify ggplot themes
library(tidyverse)
theme_set(theme_bw())
theme_update(text = element_text(family = "serif"), 
axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)))

# Correlation matrices with 1s on the diagonal
(high_cor_mat <- matrix(c(1, .9,
                         .9, 1),
           nrow = 2, ncol = 2, byrow = TRUE))
# byrow = TRUE assigns the data from left to right
(low_cor_mat <- matrix(c(1, .1,
                        .1, 1),
           nrow = 2, ncol = 2, byrow = TRUE))
```

## Run simulation using the rmvnorm function from the mvtnorm package

```{r run_sim}
 # Simulate standardized variables from two multivariate normal distributions
library(mvtnorm)
set.seed(42)
high_cor <- rmvnorm(n = 500, mean = c(0, 0), sigma = high_cor_mat) %>%
  data.frame()
set.seed(42)
low_cor <- rmvnorm(n = 500, mean = c(0, 0), sigma = low_cor_mat) %>%
  data.frame()

plot_lm <- function(data){
  
data %>%
  ggplot(aes(x = X1, y = X2)) + # Aesthetics (aes) map inputs to graph
  geom_point(alpha = 0.5) + # Alpha allows transparency in geometries
  geom_smooth(method = "lm", color = "red")
  
}
```

## Visualize the high correlation

```{r scatter}
plot_lm(high_cor)
```

## Visualize the low correlation

```{r lm}
plot_lm(low_cor)
```

## Write a p-value function for permutation testing

```{r p_value}
# This function takes as input a null distribution of a test statistic and an 
# observed test statistic.
get_p_value <- function(null, obs){
  
  p_less <- (sum(obs <= null) + 1)/(length(null) + 1)
  p_greater <- (sum(obs >= null) + 1)/(length(null) + 1)
  
  if (p_less < p_greater) {
    
    p_value <- p_less * 2
    
  } else {
    
    p_value <- p_greater * 2
    
  }
}
```

## Write the correlation permutation test function

```{r cor_test_function}
# This function takes an N X 2 dataframe as input and returns the null distribution
# of correlation values, the observed correlation value, and the probability value 
# of observing a result as extreme as the observed correlation given the 
# null distribution.
cor_perm_test <- function(data, iter = 10^4, method = "pearson"){
  
  set.seed(42)
  obs_cor <- cor(data[, 1], data[, 2], method = method)
  null_cor <- rep(NA, iter)
  
  for (i in 1:iter){
    
    index <- sample(nrow(data), replace = FALSE)
    permuted_x <- data[index, 1]
    null_cor[i] <- cor(permuted_x, data[, 2], method = method)
    
  }
  
  p_value <- get_p_value(null_cor, obs_cor)
  
  return(list(null = null_cor, obs = obs_cor, p = p_value))
  
}
```

## Run the correlation permutation tests

```{r check_sym}
# This function tests whether or not the null distribution is symmetric.
check_symmetry <- function(null_dist){
  set.seed(42)
  symm_test <- lawstat::symmetry.test(null_dist)
  
  if (symm_test$p.value > .05) {
    print(paste0("Null distribution is symmetric, p-value = ",
      symm_test$p.value))
    
  } else {
    print(paste0("Null distribution is asymmetric, p-value = ",
      symm_test$p.value, ". Consider changing the seed."))
    
 }
}
```

```{r run_cor_tests}
high_cor_test <- cor_perm_test(high_cor)
low_cor_test <- cor_perm_test(low_cor)
```

## Compute correlation permutation test p-values

```{r get_cor_p}
check_symmetry(high_cor_test$null)
high_cor_test$p
check_symmetry(low_cor_test$null)
low_cor_test$p
```

## Write a plotting function for permutation test results

```{r plot_perm}
plot_test_results <- function(data){
  
  df <- data.frame(data$null)
  names(df) <- "X1"
  
  df %>%
    ggplot(aes(x = X1)) + 
    geom_histogram(bins = sqrt(nrow(df))) +
    geom_vline(xintercept = data$obs, linetype = 2) +
    labs(x = "Null distribution of test statistic", y = "Count") +
    annotate("text", family = "serif", x = data$obs + 
               diff(range(-(abs(data$obs)), max(abs(df$X1)))) * .02, 
             y = sqrt(nrow(df)) * 2.5, label = "Observed", angle = 90,
             color = "red")
  
}
```

## High correlation test results

```{r high_test_res}
plot_test_results(high_cor_test)
```

## Low correlation test results

```{r low_test_res}
plot_test_results(low_cor_test)
```

## Set up simulation for data with mean differences

```{r mean_diff_sim}
mean_10 <- rnorm(500, 10)
mean_20 <- rnorm(500, 20)
mean_9.9 <- rnorm(500, 9.9)
groups <- rep(1:2, each = 500)

large_diff <- data.frame(my_dv = c(mean_10, mean_20),
                         my_group = groups)
small_diff <- data.frame(my_dv = c(mean_10, mean_9.9),
                         my_group = groups)
```

## Write a mean difference permutation test function

```{r mean_diff_perm}
# This function takes a dataframe, dv and grouping variable as input and returns 
# the same metrics as the cor_perm_test function. Sorry for little whitespace!
mean_diff_perm_test <- function(data, dv, group, iter = 10^4){
  set.seed(42)
  group <- enquo(group)
  dv <- enquo(dv)
  group_means <- data %>%
    group_by(!!group) %>%
    summarize(means = mean(!!dv)) %>%
    pull()
  obs_mean_diff <- group_means[1] - group_means[2]
  null_mean_diff <- rep(NA, iter)
  
  dv_data <- data %>%
    select(!!dv) %>%
    pull()
  for (i in 1:iter){
    index <- sample(nrow(data), size = nrow(data)/2, replace = FALSE)
    null_mean_diff[i] <- mean(dv_data[index]) - mean(dv_data[-index])
  }
  p_value <- get_p_value(null_mean_diff, obs_mean_diff)
  return(list(null = null_mean_diff, obs = obs_mean_diff, p = p_value))
}
```

## Run the mean difference permutation tests

```{r run_mean_diff_tests}
large_diff_test <- mean_diff_perm_test(large_diff, dv = my_dv, group = my_group)
small_diff_test <- mean_diff_perm_test(small_diff, dv = my_dv, group = my_group)
```

## Compute correlation permutation test p-values

```{r get_mean_diff_p}
check_symmetry(large_diff_test$null)
large_diff_test$p
check_symmetry(small_diff_test$null)
small_diff_test$p
```

## Large mean difference test results

```{r large_diff_res}
plot_test_results(large_diff_test)
```

## Small mean difference test results

```{r small_diff_res}
plot_test_results(small_diff_test)
```

# Now it's your turn!

Download the in-class activity and answer the questions as best you can. 

Collaborate with your peers to write code. 

I am available for programming questions.