---
title: "PSYC 7710 Lab" 
subtitle: "Lab 11 Activity"
author: "Andrew Graves, Department of Psychology, University of Virginia"
output:
  pdf_document: default
---

## Directions:

A.	Answer the following questions and save the code you used in an R script.
B.	For each question, use the **iris** dataset in base R to name the appropriate statistical test, apply the model as both a traditional test and a linear model, ensure equivalence of the two statistical models, and plot the results as a linear model.
C.	You have until the end of lab to complete.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE, warning = FALSE, message = FALSE, fig.height = 4)
```

```{r init}
library(tidyverse)
theme_set(theme_bw())
theme_update(text = element_text(family = "serif"), 
axis.title.y = element_text(margin = margin(r = 20)))

data("iris")
```

## Questions:

1. Are there differences between the means of **Petal.Width** for any of the three **Species**?

```{r 1}
# One-way analysis of variance (ANOVA)
group_means <- iris %>%
  group_by(Species) %>%
  summarize(pet_width = mean(Petal.Width))

set_mean <- group_means %>%
  filter(Species == "setosa") %>%
  select(pet_width) %>%
  pull()

vers_mean <- group_means %>%
  filter(Species == "versicolor") %>%
  select(pet_width) %>%
  pull()

virg_mean <- group_means %>%
  filter(Species == "virginica") %>%
  select(pet_width) %>%
  pull()

aov(Petal.Width ~ Species, data = iris) %>%
  summary()

lm(Petal.Width ~ Species, data = iris) %>%
  summary()

# Plot w/ treatment contrasts
iris %>%
  ggplot(aes(x = Species, y = Petal.Width)) + 
  geom_jitter(height = .1, width = 0, alpha = 0.5) +
  geom_segment(aes(x = .75, xend = 3.25, y = set_mean, yend = set_mean), color = "blue") + 
  geom_segment(aes(x = 1, xend = 2, y = set_mean, yend = vers_mean), color = "red") +
  geom_segment(aes(x = 2, xend = 3, y = set_mean, yend = virg_mean), color = "red")

# There is a difference between the means of petal width for at least one of the species.
```

2. Is **Sepal.Width** for the **Species** *virginica* less than 3 cm?

```{r 2}
# One sample t-test
one_samp_data <- iris %>%
  filter(Species == "virginica") %>%
  select(Species, Sepal.Width)

t.test(one_samp_data$Sepal.Width, mu = 3)

lm(one_samp_data$Sepal.Width - 3 ~ 1) %>%
  summary()

one_samp_data %>%
  ggplot(aes(x = Species, y = Sepal.Width)) + 
  geom_jitter(height = .1, width = 0, alpha = 0.5) +
  geom_hline(yintercept = 3, lty = 2) + 
  geom_segment(aes(x = .75, xend = 1.25, 
              y = mean(one_samp_data$Sepal.Width), 
              yend = mean(one_samp_data$Sepal.Width)), color = "blue")

# There is little to no evidence that virginica sepal width is less than 3 cm.
```

3. Is there a difference between the means of **Petal.Length** for the **Species** *versicolor* and *virginica*?

```{r 3}
# Independent samples t-test
ind_samp_data <- iris %>%
  filter(Species %in% c("versicolor", "virginica")) %>%
  select(Species, Petal.Length)

ind_samp_data_vers <- ind_samp_data %>%
  filter(Species == "versicolor") %>%
  select(Petal.Length) %>%
  pull()

ind_samp_data_virg <- ind_samp_data %>%
  filter(Species == "virginica") %>%
  select(Petal.Length) %>%
  pull()

t.test(ind_samp_data_virg, ind_samp_data_vers, var.equal = TRUE)

lm(Petal.Length ~ Species, data = ind_samp_data) %>%
  summary()

ind_samp_data %>%
  ggplot(aes(x = Species, y = Petal.Length, group = 1)) + 
  geom_jitter(height = .1, width = 0, alpha = 0.5) +
  stat_summary(fun.y = mean, geom = "line", color = "red") +
  geom_segment(aes(x = .75, xend = 1.25, y = mean(ind_samp_data_vers), 
               yend = mean(ind_samp_data_vers)), color = "blue")

# The virginica petal length is larger than the versicolor petal length. 
```

4. Across all the flowers, are the petals shorter than the sepals?

```{r 4}
# Paired samples t-test
t.test(iris$Petal.Length, iris$Sepal.Length, paired = TRUE)

lm(Petal.Length - Sepal.Length ~ 1, data = iris) %>%
  summary()

add_length_diff <- iris %>%
  mutate(`Length Difference` = Petal.Length - Sepal.Length,
         `All Flowers` = "All Flowers")

add_length_diff %>%
  ggplot(aes(x = `All Flowers`, y = `Length Difference`)) + 
  geom_jitter(height = .1, width = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, lty = 2) + 
  geom_segment(aes(x = .75, xend = 1.25, y = mean(add_length_diff$`Length Difference`), 
  yend = mean(add_length_diff$`Length Difference`)), color = "blue") + 
  labs(x = "")

# Petals are shorter than the sepals. 
```
